---
title: "New try"
author: "O.M."
date: "2023-11-02"
output:
  word_document: default
  html_document: default
---

```{r}
library(tidyverse)
library(readr)
library(tidyr)
library(stringr)
library(AnnotationHub)
library(RColorBrewer)
library(readxl)
library(DESeq2)
library(pheatmap)
library(DEGreport)
library(tximport)
library(ggrepel)
library(knitr)
library(biomaRt) 
library(ensembldb) 
library(plotly)
library(fdrtool)
library(VennDetail)
library(enrichR)
library(pathview)
library(ggstatsplot)
library(vegan)
```



```{r}
df<-read.csv("count_matrix_join_f.csv")%>%select(2, 3, 5, 7, 9, 11, 13, 15, 17)
row.names(df)<-df$X
df$X<-NULL
meta<-read_excel("metadata_at2.xlsx")
rownames(meta)<-meta$sample_id
sample_min<-min(colSums(df))
set.seed(2023)
df<-df%>%t()%>%rrarefy(sample_min)%>%t()%>%as.data.frame()


```
### Biomart annotations protein coding genes filtering and zero removal {.tabset}
```{r Biomart annotations protein coding genes filtering and zero removal}
ensembl <- useEnsembl(biomart = "ensembl", dataset = "rnorvegicus_gene_ensembl")

annotations <- getBM(attributes = c("ensembl_gene_id", "gene_biotype", "external_gene_name", "description", "entrezgene_id", "ensembl_transcript_id"), mart = ensembl)

protein_coding_annotations <- annotations[annotations$gene_biotype == "protein_coding", ]

protein_coding_gene_ids <- protein_coding_annotations$ensembl_transcript_id

filtered_counts <- df[row.names(df) %in% protein_coding_gene_ids, ]
#filtered_counts<-filtered_counts%>%dplyr::filter(abdominal1>=8, abdominal2>=8, abdominal4>=8, abdominal5>=8, thoracic1>=8, thoracic2>=8, mesenterium1>=8, mesenterium2>=8)


```






### DDS object creation {.tabset}
```{r DDS object creation }

    dds <- DESeqDataSetFromMatrix(countData = filtered_counts, colData = meta, design = ~location)
    dds$location <- as.factor(dds$location)
    keep <- rowMeans(counts(dds)) >= 60
    dds <- dds[keep, ]

ds<-DESeq(dds, test="LRT", reduced = ~1)
```


## PCA - location
```{r}
rld <- vst(ds)
rld_mat <- assay(rld)
plt_pca<-plotPCA(rld, intgroup = c("location")) + geom_label_repel(aes(label = name), label.size = 0.01, show.legend = F, box.padding = 0.12, size = 2.5) + theme_bw()+scale_color_manual(values=c( "#FFF019", "#107A9E", "#736F33" ))+theme(legend.position = "none", axis.title =  element_text(size=4.7))

ggsave("plt_pca.tiff",plt_pca, width = 8.9,
  height = 6.6,
  dpi = 1200, units = "cm")
```

### lfc shrinkage {.tabset}
```{r lfc shrinkage, message=FALSE}
resultsNames(ds)
Deseq_results <- results(ds, contrast=c("location","thoracic", "abdominal"), alpha = 0.05)
Deseq_results_cor <- fdrtool(Deseq_results$stat, statistic= "normal", plot = T) 
Deseq_results <- lfcShrink(ds, coef="location_thoracic_vs_abdominal", res=Deseq_results, svalue = T)
Deseq_results$padj  <- p.adjust(Deseq_results_cor$pval, method = "BH")

```


```{r extracting significant genes}
dex_results <- Deseq_results %>% as.data.frame() %>% 
  rownames_to_column(var = "gene") %>% 
  as_tibble() %>% 
  left_join(protein_coding_annotations, by = c("gene" = "ensembl_transcript_id"))

dex_results_significant <- dplyr::filter(dex_results, padj < 0.01) %>%  
                        dplyr::filter(abs(log2FoldChange) > 1.5) %>% dplyr::filter(gene_biotype=="protein_coding"&external_gene_name!="")

dex_results_significant%>%arrange(desc(log2FoldChange))%>%select(-11)%>%unique()%>%writexl::write_xlsx("All differentially expressed genes.xlsx")
```

### Volcano plot {.tabset}
```{r, fig.width=8}

p<-dex_results%>%mutate(Significance=ifelse(log2FoldChange>1.5&padj>0.05, "No significant difference", ifelse(log2FoldChange<1.5&padj<0.05&log2FoldChange>(-1.5)| log2FoldChange>(-1.5)&padj<0.05&log2FoldChange<(1.5), "No significant difference", ifelse(log2FoldChange<1.5&padj>0.05&log2FoldChange>(-1.5)| log2FoldChange>(-1.5)&padj>0.05&log2FoldChange<(1.5), "No difference", ifelse(log2FoldChange>1.5&padj<0.05, "Expression higher in thoracic", ifelse(log2FoldChange<(-1.5)&padj<0.05, "Expression higher in abdominal", "No significant difference"))))))



p2<-ggplot(p, aes(x=log2FoldChange, y=-log10(padj),label=external_gene_name))+geom_point(aes(color=Significance), size=0.33)+theme_light()+scale_x_continuous(limits = c(-6, 6))+geom_vline(xintercept=c(-1.5, 1.5), col="red") + geom_hline(yintercept=-log10(0.05), col="red")+labs(title = "")+scale_y_continuous(limits = c(-0.4,12.4))+theme(legend.text=element_text(size=4.11), legend.title = element_text(size = 4.11), legend.spacing.x = unit(-0.03, 'cm'),   legend.spacing.y = unit(0.01, "cm"))+scale_color_manual(values=c( "#FFF019", "#107A9E", "#000000","#736F33"))

ggplotly(p2)

ggsave("volcano.tiff",p2, width = 11.5,
  height = 6.8,
  dpi = 1200, units = "cm")
```



## enrichment analysis{.tabset}
### GO_Molecular_Function_2021{.tabset}
```{r GO_Molecular_Function_2021, fig.height=10, fig.width=12}
setEnrichrSite("Enrichr")

dbs <- listEnrichrDbs()
dbs <- c("GO_Biological_Process_2021", "HDSigDB_Mouse_2021", "Mouse_Gene_Atlas", "GO_Molecular_Function_2021", "KEGG_2021_Human")

aorta_up <- dex_results_significant %>%as.data.frame()%>%dplyr::filter(svalue < 0.05 & log2FoldChange > 1.5)
mesenterium_up <- dex_results_significant %>%as.data.frame()%>%dplyr::filter(svalue < 0.05 & log2FoldChange < -1.5)

ao_up <- enrichr(aorta_up$external_gene_name, dbs)
mes_up <- enrichr(mesenterium_up$external_gene_name, dbs)


ao_up_sig <- ao_up$GO_Biological_Process_2021 %>% 
              dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))


mes_up_sig <- mes_up$GO_Biological_Process_2021 %>% 
               dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))


dataframe_to_enrichr <- function(dataframe){
  tmp = dataframe %>% tidyr::separate("Overlap", sep = "/", into = c("Gene_number", "max"))
  tmp$Gene_number = as.numeric(tmp$Gene_number)
  tmp = tmp %>% arrange(desc(Gene_number))
}



ao_up_sig <- dataframe_to_enrichr(ao_up_sig)



mes_up_sig <- dataframe_to_enrichr(mes_up_sig)

  
ggplot(ao_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Thoracic")+theme_ggstatsplot()



ggplot(mes_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Abdominal")+theme_ggstatsplot()
  
```

```{r}

```

### KEGG_2021_Human{.tabset}
```{r KEGG_2021_Human2, fig.height=10, fig.width=12}
ao_up_sig <- ao_up$KEGG_2021_Human %>% 
              dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))


mes_up_sig <- mes_up$KEGG_2021_Human %>% 
               dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))



ao_up_sig <- dataframe_to_enrichr(ao_up_sig)



mes_up_sig <- dataframe_to_enrichr(mes_up_sig)



ggplot(ao_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Aorta")


ggplot(mes_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Mesenterium")

```


### KEGG_2021_Human{.tabset}
```{r KEGG_2021_Human, fig.height=10, fig.width=12}
ao_up_sig <- ao_up$HDSigDB_Mouse_2021 %>% 
              dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))


mes_up_sig <- mes_up$HDSigDB_Mouse_2021 %>% 
               dplyr::filter(P.value < 0.01) %>% arrange(desc(Combined.Score))



ao_up_sig <- dataframe_to_enrichr(ao_up_sig)



mes_up_sig <- dataframe_to_enrichr(mes_up_sig)



ggplot(ao_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Aorta")


ggplot(mes_up_sig, aes(x = reorder(Term, Gene_number),
                           y = Gene_number,
                           fill = -log(P.value, base = 10))) +
  geom_bar(stat = "identity", width = 0.5, alpha = 0.8) + coord_flip()+labs(title = "Mesenterium")

```

 
## pathwiev analysis{.tabset}
### contractility  {.tabset}
```{r contractility, fig.height=10, fig.width=12}
logFC <- dex_results_significant$log2FoldChange
names(logFC) <- dex_results_significant$entrezgene_id

pathview(gene.data = logFC, 
         pathway.id = "rno04270", 
         species = "rno", 
         kegg.native = T)

```

```{r}

```

### PPARy{.tabset}
```{r PPARy, fig.height=10, fig.width=12}
pathview(gene.data = logFC, 
         pathway.id = "rno03320", 
         species = "rno", 
         kegg.native = T)

```

```{r}

```

### gap junction{.tabset}
```{r gap junction, fig.height=10, fig.width=12}
pathview(gene.data = logFC, 
         pathway.id = "rno04540", 
         species = "rno", 
         kegg.native = T)

```

```{r}

```

### ach signaling {.tabset}
```{r ach signaling, fig.height=10, fig.width=12}
pathview(gene.data = logFC, 
         pathway.id = "rno04725", 
         species = "rno", 
         kegg.native = T)

```

```{r}

```

### TRPV{.tabset}
```{r TRPV, fig.height=10, fig.width=12}
pathview(gene.data = logFC, 
         pathway.id = "rno04750", 
         species = "rno", 
         kegg.native = T)

```




